The role of lithic bipolar technology in Western Iberia’s Upper Paleolithic: the case of Vale Boi (southern Portugal)

### 2018-05-13 analysis/paper/paper.Rmd
advances in chi-square functions

In scaled pieces, metrics are a difficult set of variables to clearly show as results. These attributes are impacted by both the initial blank size as well as by the functional use and reduction. Consideranging the typological axes (see Table \@ref(tab:tech_attributes)) for definition) length ranging from `r min(techno_data$AxisLength, na.rm = TRUE)` mm to `r max(techno_data$AxisLength, na.rm = TRUE)` mm, width ranging from `r min(techno_data$AxisWidth, na.rm = TRUE)` mm to `r max(techno_data$AxisWidth, na.rm = TRUE)` mm and thickness ranging from `r min(techno_data$Thickness, na.rm = TRUE)` mm to `r max(techno_data$Thickness, na.rm = TRUE)` mm. In general, mean values tend to be similiar between raw materials, with some differences occuring within the Proto-Solutrean and Magdalenian assemblages, most certainly as a result of the small samples analysed for each of these periods (Figure \@ref(fig:meanplots)).

```{r meanplots, echo = FALSE, fig.cap="Barplots of means for Length, Width and Thickness of scaled pieces by raw material and across the four chronological phases. Error bars represent standard deviations."}

length <- mean_plot(techno_data, "Chronology", "AxisLength", "RawMaterial")
width <- mean_plot(techno_data, "Chronology", "AxisWidth", "RawMaterial")
thickness <- mean_plot(techno_data, "Chronology", "Thickness", "RawMaterial")

meanplots <- ggarrange(length, width, thickness, ncol = 1, nrow = 3, labels = c("A", "B", "C"), common.legend = TRUE, legend = "right")

@@

>>>>

As in other classes of stone tools, scaled pieces' metric attributes are impacted by both the initial blank size as well as by the intensity of their use. Specifically, because some scaled pieces are used in multiple axes, direct comparisons for length and width of the typological axes (see Table \@ref(tab:tech_attributes)) for definition) cannot be straighforwardly performed. Since the majority of the analyzed pieces presented a rectangular outline, we use the area of a rectangle (i.e. Length x Width) as an approximation for the overall dimensions of the artifacts. Area calculations revealed a maximum of `r round(max(techno_data$Area, na.rm = TRUE), digits = 2)` mm and a minimum of `r round(max(techno_data$Area, na.rm = TRUE), digits = 2)` mm. For thickness the maximum is `r round(max(techno_data$Thickness, na.rm = TRUE), digits = 2)` mm and the minimum is `r round(min(techno_data$Thickness, na.rm = TRUE), digits = 2)` mm. In general, mean values tend to be similiar between raw materials, with some differences occuring within the Proto-Solutrean and Magdalenian assemblages, most certainly as a result of the small samples analysed for each of these periods (Figure \@ref(fig:meanplots)). Across techno-complexes, however, no significant statistical differences were detected when using an ANOVA test for both Area (`r anova_apa(aov(Area~Chronology, techno_data), "Chronology")` and  Thickness (`r anova_apa(aov(Thickness~Chronology, techno_data), "Chronology")`).

When plotted against the number of damaged platforms, results for the Area variable reveal that pieces with four damaged platforms are amongst the largest, and that there is not a visible difference between the thickness of the pieces comprising each group.

```{r meanplots, echo = FALSE, fig.cap="Barplots of means for Area (Length x Width) and Thickness of scaled pieces by raw material and across the four chronological phases. Error bars represent standard deviations."}

techno_data <- techno_data %>% 
  mutate(Area = AxisWidth*AxisLength)

area <- mean_plot(techno_data, "Chronology", "Area", "RawMaterial")
thickness <- mean_plot(techno_data, "Chronology", "Thickness", "RawMaterial")

meanplots <- ggarrange(area, thickness, ncol = 1, nrow = 2, common.legend = TRUE, legend = "right")

@@

```{r plotbyplatforms, echo = FALSE, fig.caption = ""}

area_by_platforms <- box_plot(techno_data, "DamagedPlatforms", "Area", "", "Area (Length * Width)")
thickness_by_platforms <- box_plot(techno_data, "DamagedPlatforms", "Thickness", "Number of Damaged Platforms", "Thickness (mm)")

boxplots <- ggarrange(area_by_platforms, thickness_by_platforms, ncol = 1, nrow = 2, common.legend = TRUE, legend = "right")

boxplots

With few exceptions, most of the morpho-functional attributes show a higher degree of variability than technological attributes, and no significant patterns could be clearly noticed. In fact, only two attributes were relatively constant across all occupations: most damaged platforms present less than 45º angles; and, a large percentage of pieces with two damaged platforms present bifacial scars on one of the platforms and unifacial scars on the opposite.

On a closer inspection, we find that most of the differences within the morpho-functional variables occur between raw materials rather than between chronologies.

```{r }

# Filter attributes
morpho_data <- all_data$morpho_data
morpho_data <- morpho_data %>% 
  filter(DamagedPlatforms == 2)
  select(RawMaterial, Chronology,
         ScarDistribution_PlatformA, ScarDistribution_PlatformB,
         ScarArrangement_PlatformA,ScarArrangement_PlatformB,
         ScarExtention_PlatformA,ScarExtention_PlatformB,
         ScarFacialDistribution_PlatformA, ScarFacialDistribution_PlatformB,
         ScarEdgeDelineation_PlatformA, ScarEdgeDelineation_PlatformB,
         Angle_PlatformA, Angle_PlatformB)

var_list <- c("DamageDegree", "ScarShape", "ScarDistribution", "ScarArrangement", "ScarExtension", "ScarFacialDistribution")

# Filter by chronology
filter <- split(morpho_data_2plat, morpho_data$Chronology)

# Recode based on 10% frequency rule
condense(morpho_data)

# Join attributes
morpho_data_2plat <- morpho_data %>% 
  mutate(DamageDegree = str_c(DamageDegree_PlatformA,DamageDegree_PlatformB, sep = "/")) %>% 
  mutate(ScarShape = str_c(ScarShape_PlatformA,ScarShape_PlatformB, sep = "/")) %>%
  mutate(ScarDistribution = str_c(ScarDistribution_PlatformA,ScarDistribution_PlatformB, sep = "/")) %>%
  mutate(ScarArrangement = str_c(ScarArrangement_PlatformA,ScarArrangement_PlatformB, sep = "/")) %>%
  mutate(ScarExtension = str_c(ScarExtention_PlatformA,ScarExtention_PlatformB, sep = "/")) %>%
  mutate(ScarFacialDistribution = str_c(ScarFacialDistribution_PlatformA,ScarFacialDistribution_PlatformB, sep = "/")) %>%
  mutate(Scar_EdgeDelineation = str_c(ScarEdgeDelineation_PlatformA,ScarEdgeDelineation_PlatformB, sep = "/")) %>%
  mutate(Angle = str_c(Angle_PlatformA,Angle_PlatformB, sep = "/"))
